Concurrent Programming with CXL

### 2020-05-18 book/paper.tex
Partially updated book to use process 0 and 1 rather than 1 and 2

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.  The program defines a
method \texttt{up} that takes a parameter called \texttt{self}.
It also initializes two shared variables \texttt{count} and \texttt{done}.
Variable \texttt{done} is initialized with a dictionary that maps both 1 and 2
to \texttt{False}.
The \texttt{up()} method increments \texttt{count} and sets \texttt{done[self]}
to \texttt{True}.
Next the program spawns two processes, each invoking \texttt{up}.
Afterward there are three processes---after
all, there was already one, and it spawned two new ones.  We will call
the one that was already there the \emph{main process}.

@@

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved in unspecified and often random ways.
In fact, CXL will try \emph{all possible interleavings} of the processes
executing machine instructions.
And thus, the following is a possible interleaving of
the processes 1 and 2:
\begin{enumerate}
\item Process 1 loads the value of \texttt{count}, which is 0;
\item Process 2 loads the value of \texttt{count}, which is still 0;
\item Process 2 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 1 sets \texttt{upDone} to \texttt{True};
\item Process 2 sets \texttt{upDone} to \texttt{True}.
\end{enumerate}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def up(self):
   0 Jump 12
   1 Frame ('up', 'Up.cxl', 1, 5) ('self', 'Up.cxl', 1, 8) 11
Up.cxl:2     count = count + 1;
   2 Constant (1, 'Up.cxl', 2, 21)
   3 Load ('count', 'Up.cxl', 2, 13)
   4 Nary ('+', 'Up.cxl', 2, 19) 2
   5 Name ('count', 'Up.cxl', 2, 5)
   6 Store 1
Up.cxl:3     done[self] = True;
   7 Constant (True, 'Up.cxl', 3, 18)
   8 LoadVar ('self', 'Up.cxl', 3, 10)
   9 Name ('done', 'Up.cxl', 3, 5)
   10 Store 2
   11 Return
Up.cxl:5 count = 0;
   12 Constant (0, 'Up.cxl', 5, 9)
   13 Name ('count', 'Up.cxl', 5, 1)
   14 Store 1
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
Assertion failed ('assert', '<stdin>', 12, 1) 1

@@

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{dict\{ .name: "\_\_main\_\_", .tag: () \}}, pc 0,
register \texttt{dict\{\}}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

If we look at the middle three rows, we see that:
\begin{enumerate}
\item Process 1 executed instructions 1 through 5, loading the value of
\texttt{count} but stopping just before storing 1 into \texttt{count};
\item Process 2 executed instructions 1 through 10, storing 1 into
into \texttt{count} and storing \texttt{True} into \texttt{done[2]};
\item Process 1 continues execution, storing value 1 into \texttt{count}
and storing \texttt{True} into \texttt{done[1]}.
\end{enumerate}

>>>>

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.  The program defines a
method \texttt{process} that takes a parameter called \texttt{self}.
It also initializes two shared variables \texttt{count} and \texttt{done}.
Variable \texttt{done} is initialized with a dictionary that maps both 1 and 2
to \texttt{False}.
The \texttt{process()} method increments \texttt{count} and sets \texttt{done[self]}
to \texttt{True}.
Next the program spawns two processes, each invoking \texttt{process}.
Afterward there are three processes---after
all, there was already one, and it spawned two new ones.  We will call
the one that was already there the \emph{main process}.

@@

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved in unspecified and often random ways.
In fact, CXL will try \emph{all possible interleavings} of the processes
executing machine instructions.
And thus, the following is a possible interleaving of
the processes 0 and 1:
\begin{enumerate}
\item Process 0 loads the value of \texttt{count}, which is 0;
\item Process 1 loads the value of \texttt{count}, which is still 0;
\item Process 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 sets \texttt{upDone} to \texttt{True};
\item Process 1 sets \texttt{upDone} to \texttt{True}.
\end{enumerate}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def process(self):
   0 Jump 12
   1 Frame ('process', 'Up.cxl', 1, 5) ('self', 'Up.cxl', 1, 13) 11
Up.cxl:2     count = count + 1;
   2 Constant (1, 'Up.cxl', 2, 21)
   3 Load ('count', 'Up.cxl', 2, 13)
   4 Nary ('+', 'Up.cxl', 2, 19) 2
   5 Name ('count', 'Up.cxl', 2, 5)
   6 Store 1
Up.cxl:3     done[self] = True;
   7 Constant (True, 'Up.cxl', 3, 18)
   8 LoadVar ('self', 'Up.cxl', 3, 10)
   9 Name ('done', 'Up.cxl', 3, 5)
   10 Store 2
   11 Return
Up.cxl:5 count = 0;
   12 Constant (0, 'Up.cxl', 5, 9)
   13 Name ('count', 'Up.cxl', 5, 1)
   14 Store 1
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
Assertion failed ('assert', 'Up.cxl', 12, 1) 1

@@

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty tuple, \texttt{()}.
The context bag has an initial context in it with name tag
\texttt{dict\{ .name: "\_\_main\_\_", .tag: () \}}, pc 0,
register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

If we look at the middle three rows, we see that:
\begin{enumerate}
\item Process 0 executed instructions 1 through 5, loading the value of
\texttt{count} but stopping just before storing 1 into \texttt{count};
\item Process 1 executed instructions 1 through 10, storing 1 into
into \texttt{count} and storing \texttt{True} into \texttt{done[2]};
\item Process 0 continues execution, storing value 1 into \texttt{count}
and storing \texttt{True} into \texttt{done[1]}.
\end{enumerate}
